---
title: "Insight-Oriented Data Projects Resources"
title-block-banner: true
format: 
 html: 
  toc: true
execute: 
  echo: false
  warning: false
---

```{r}
#| label: load-libraries


library(DT)

```

# Insight-Oriented Data Projects Resources

An insight-oriented data project starts with a clear purpose: to discover, understand, explain, or solve business problems. Significant effort is focused on defining the problem, formulating the question you want to answer, and designing an analysis workflow to provide insights for resolving the question. Data are collected, analyzed, and interpreted systematically to address specific problems and assist with making informed decisions.



```{r}
#| column: page-right

# Create a matrix

dfg <- matrix(c('<a href="https://www.designative.info/2021/03/27/strategy-and-stakeholder-management/">Strategy and Stakeholder Management</a>', 
                '<a href="https://libguides.library.kent.edu/SPSS">Tutorials for working with data</a>',
                '<a href="https://online.hbs.edu/blog/post/data-storytelling">Data Storytelling: How to effectively tell a story with data</a>',
                '<a href="https://www.youtube.com/@storytellingwithdata">storytelling with data</a>',
                '<a href="https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6322175/">Formulation of Research Question – Stepwise Approach</a>', 
                '<a href="https://hbr.org/2015/03/relearning-the-art-of-asking-questions?referral=00060">Relearning the Art of Asking Questions</a>', 
                '<a href="https://www.open.edu/openlearncreate/mod/oucontent/view.php?id=80245">Problem Definition</a>' ,
                '<a href="https://www.selectstar.com/blog/why-youre-closer-to-data-documentation-than-you-think">Why You’re Closer to Data Documentation Than You Think</a>', 
                
                
                "Stakeholder Management skills that help you identify people, groups, or organizations that could impact or be impacted by projects.",
                "Written and illustrated tutorials for the statistical software SPSS.",
                "How to communicate the story it tells in a clear, compelling manner—a skill called data storytelling.",
                "Watch, learn, and become a better data storyteller through our videos and webinars.",
                "This paper discusses the process of formulating RQ with a stepwise approach.",
                "The four questions you need to solve different types of problems.",
                "The Problem Definition Tool guides you to define a problem analytically, exploring it from different angles and helping clarify your priorities by focusing on key critical issues.",
                "Documentation isn’t a separate task from what you’ve already been doing, so you’re closer to well-documented data than you think!",
                
                'Stakeholder management',
                'Data Exploration',
                'Tell the data story',
                'Tell the data story',
                'Research Question',
                'Asking Questions',
                'Problem Definition',
                'Data documentation',
               
                'Article',
                'Tutorial',
                'Article',
                'Video',
                'Article',
                'Article',
                'Course',
                'Blog article'
             ),
              
              ncol = 4)
colnames(dfg) <- c('Title', 'Description', 'Topic', 'Type')

# Create a datatable
datatable(dfg, escape = FALSE)  # Set escape = FALSE to render HTML content




```

