---
title: "Quarto Tools"
title-block-banner: true
format: 
 html: 
  toc: true
execute: 
  echo: false
  warning: false
---

```{r}
#| label: load-libraries


library(DT)
library(shiny)

```

# Quarto Tools

[Quarto](https://quarto.org/) is an open-source scientific and technical publishing system with support for multiple languages and environments. It offers a wide range of functionalities, making it versatile for various tasks. There are many resources on the web to learn about Quarto. Here, I have collected some of them that I find particularly helpful for my Quarto code, blog, presentations, and workflow.

If you want to start your journey with Quarto, I recommend the [Quarto documentation](https://quarto.org/), which is the best place to learn everything you might want to know about Quarto.

<br>

```{r}
#| column: page-right

qt <- matrix(c('#quartotip',
               '#buildblog',
               '#documenting #r #reporting', 
               '#documenting #r #reporting',
               
               "A website that highlights a tip for Quarto every day for a month.",
               " An in-depth guide that shows how to build a blog with Quarto. A great resource that helps you start your own blog.",
               "Generate documents, presentations and more. Produce HTML, PDF, MS Word, reveal.js, MS Powerpoint, Beamer, websites, blogs, books",
               "Data reporting and communication",
               
               
                "tips/tweets",
                'tutorial',
                'html',
                'pdf',
               
               
               
               '<a href="https://mine-cetinkaya-rundel.github.io/quarto-tip-a-day/">A Quarto tip a day</a>',
                '<a href="https://albert-rapp.de/posts/13_quarto_blog_writing_guide/13_quarto_blog_writing_guide.html">The ultimate guide to starting a Quarto blog</a>',
                '<a href="https://rstudio.github.io/cheatsheets/html/quarto.html">Publish and Share with Quarto :: Cheatsheet</a>',
               '<a href="https://res.cloudinary.com/dyd911kmh/image/upload/v1676540721/Marketing/Blog/Quarto_Cheat_Sheet.pdf">Quarto Cheat Sheet</a>'),
             
             ncol = 4)

colnames(qt) <- c('Category', 'Description', 'Type', 'Details')


datatable(qt, escape = FALSE) 
          
          


```
