---
title: "Quarto Tools"
title-block-banner: true
format: 
 html: 
  toc: true
execute: 
  echo: false
  warning: false
---

```{r}
#| label: load-libraries


library(DT)
library(shiny)

```

# Quarto Tools

[Quarto](https://quarto.org/) is an open-source scientific and technical publishing system with support for multiple languages and environments. It offers a wide range of functionalities, making it versatile for various tasks. There are many resources on the web to learn about Quarto. Here, I have collected some of them that I find particularly helpful for my Quarto code, blog, presentations, and workflow.

If you want to start your journey with Quarto, I recommend the [Quarto documentation](https://quarto.org/), which is the best place to learn everything you might want to know about Quarto.

<br>

```{r}
#| column: page-right

qt <- matrix(c('#documenting #r #reporting', 
               '#documenting #r #reporting',
               "Generate documents, presentations and more. Produce HTML, PDF, MS Word, reveal.js, MS Powerpoint, Beamer, websites, blogs, books",
               "Data reporting and communication",
                'NA',
                'NA',
                '<a href="https://rstudio.github.io/cheatsheets/html/quarto.html">Publish and Share with Quarto :: Cheatsheet</a>',
               '<a href="https://res.cloudinary.com/dyd911kmh/image/upload/v1676540721/Marketing/Blog/Quarto_Cheat_Sheet.pdf">Quarto Cheat Sheet</a>'), ncol = 4)

colnames(qt) <- c('Category', 'Description', 'Code', 'Details')


datatable(qt, escape = FALSE) 
          
          


```
